---
title: "Assignment 4: Data Wrangling"
author: "Kelly Davidson"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
#This code chunk will tidy your knit PDF files, wrapping long code lines
#For it to work, the "formatR" package needs to be installed

install.packages('formatR')
library(formatR)
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE)
```

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Wrangling 

## Directions
1.  Rename this file `<FirstLast>_A03_DataExploration.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.

The completed exercise is due on Friday, Oct7th @ 5:00pm.

## Set up your session 
1. Check your working directory, load the `tidyverse` and `lubridate` packages, and upload all four raw data files associated with the EPA Air dataset, being sure to set string columns to be read in a factors. See the README file for the EPA air datasets for more information (especially if you have not worked with air quality data previously). 

2. Explore the dimensions, column names, and structure of the datasets. 

```{r, message = FALSE}
#1. Checking working directory, loading 'tidyverse,' 'lubridate,' and 'dplyr' packages, and uploading 4 EPA air datasets
getwd()
library(tidyverse)
library(lubridate)
library(dplyr)

EPAair_PM25_2019 <- read.csv("./Data/Raw/EPAair_PM25_NC2019_raw.csv" , stringsAsFactors = TRUE)
EPAair_PM25_2018 <- read.csv("./Data/Raw/EPAair_PM25_NC2018_raw.csv" , stringsAsFactors = TRUE)
EPAair_O3_2019 <- read.csv("./Data/Raw/EPAair_O3_NC2019_raw.csv" , stringsAsFactors = TRUE)
EPAair_O3_2018 <- read.csv("./Data/Raw/EPAair_O3_NC2018_raw.csv" , stringsAsFactors = TRUE)

#2. Exploring dimensions, column names, and structure of each dataset
dim(EPAair_PM25_2019)
colnames(EPAair_PM25_2019)
str(EPAair_PM25_2019)

dim(EPAair_PM25_2018)
colnames(EPAair_PM25_2018)
str(EPAair_PM25_2018)

dim(EPAair_O3_2019)
colnames(EPAair_O3_2019)
str(EPAair_O3_2019)

dim(EPAair_O3_2018)
colnames(EPAair_O3_2018)
str(EPAair_O3_2018)

```

## Wrangle individual datasets to create processed files.
3. Change date to date
4. Select the following columns: Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, SITE_LONGITUDE
5. For the PM2.5 datasets, fill all cells in AQS_PARAMETER_DESC with "PM2.5" (all cells in this column should be identical).
6. Save all four processed datasets in the Processed folder. Use the same file names as the raw files but replace "raw" with "processed".

```{r}
#3. Changing the "Date" columns to dates instead of factors for each dataset
EPAair_PM25_2019$Date <- as.Date(EPAair_PM25_2019$Date, format = "%m/%d/%Y")
EPAair_PM25_2018$Date <- as.Date(EPAair_PM25_2018$Date, format = "%m/%d/%Y")
EPAair_O3_2019$Date <- as.Date(EPAair_O3_2019$Date, format = "%m/%d/%Y")
EPAair_O3_2018$Date <- as.Date(EPAair_O3_2018$Date, format = "%m/%d/%Y")

#4. Selecting the columns Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY, SITE_LATITUDE, and SITE_LONGITUDE for each dataset
EPAair_PM25_2019 <-
  EPAair_PM25_2019 %>% 
  select(Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY:SITE_LONGITUDE)

EPAair_PM25_2018 <-
  EPAair_PM25_2018 %>% 
  select(Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY:SITE_LONGITUDE)

EPAair_O3_2019 <-
  EPAair_O3_2019 %>% 
  select(Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY:SITE_LONGITUDE)

EPAair_O3_2018 <-
  EPAair_O3_2018 %>% 
  select(Date, DAILY_AQI_VALUE, Site.Name, AQS_PARAMETER_DESC, COUNTY:SITE_LONGITUDE)

#5. For the two PM2.5 datasets (2018 & 2019), filling in all cells within the "AQS_PARAMETER_DESC" column with "PM2.5"
EPAair_PM25_2019 <-
  EPAair_PM25_2019 %>% 
  mutate(AQS_PARAMETER_DESC = "PM2.5")

EPAair_PM25_2018 <-
  EPAair_PM25_2018 %>% 
  mutate(AQS_PARAMETER_DESC = "PM2.5")

#6. Saving all 4 processed datasets in the 'Processed' folder
write.csv(EPAair_PM25_2019, row.names = FALSE,
  file = "./Data/Processed/EPAair_PM25_NC2019_processed.csv")
write.csv(EPAair_PM25_2018, row.names = FALSE,
  file = "./Data/Processed/EPAair_PM25_NC2018_processed.csv")
write.csv(EPAair_O3_2019, row.names = FALSE,
  file = "./Data/Processed/EPAair_O3_NC2019_processed.csv")
write.csv(EPAair_O3_2018, row.names = FALSE,
  file = "./Data/Processed/EPAair_O3_NC2018_processed.csv")

```


## Combine datasets
7. Combine the four datasets with `rbind`. Make sure your column names are identical prior to running this code. 
8. Wrangle your new dataset with a pipe function (%>%) so that it fills the following conditions: 

* Include all sites that the four data frames have in common: "Linville Falls", "Durham Armory", "Leggett", "Hattie Avenue", "Clemmons Middle", "Mendenhall School", "Frying Pan Mountain", "West Johnston Co.", "Garinger High School", "Castle Hayne", "Pitt Agri. Center", "Bryson City", "Millbrook School" (the function `intersect` can figure out common factor levels)
* Some sites have multiple measurements per day. Use the split-apply-combine strategy to generate daily means: group by date, site, aqs parameter, and county. Take the mean of the AQI value, latitude, and longitude.
* Add columns for "Month" and "Year" by parsing your "Date" column (hint: `lubridate` package)
* Hint: the dimensions of this dataset should be 14,752 x 9.

9. Spread your datasets such that AQI values for ozone and PM2.5 are in separate columns. Each location on a specific date should now occupy only one row.
10. Call up the dimensions of your new tidy dataset.
11. Save your processed dataset with the following file name: "EPAair_O3_PM25_NC1819_Processed.csv"

```{r}
#7. Combining the 4 EPA air datasets into one using 'rbind' function
EPAair_18_19 <- rbind(EPAair_PM25_2019, EPAair_PM25_2018, EPAair_O3_2019, EPAair_O3_2018)

#8. Wrangling new data set using a pipe function
EPAair_18_19 <-
  EPAair_18_19 %>% 
  filter(Site.Name %in% c("Linville Falls", "Durham Armory", "Leggett", "Hattie Avenue", 
      "Clemmons Middle", "Mendenhall School", "Frying Pan Mountain", "West Johnston Co.",
      "Garinger High School", "Castle Hayne", "Pitt Agri. Center", "Bryson City", "Millbrook School")
      ) %>% #filtering the 'Site.Name' column to include only the listed sites
  group_by(Date, Site.Name, AQS_PARAMETER_DESC, COUNTY
           ) %>% #generating daily means and grouping by date, site, AQS parameter, & county
      summarise(meanAQI = mean(DAILY_AQI_VALUE),
            meanLatitude = mean(SITE_LATITUDE),
            meanLongitude = mean(SITE_LONGITUDE)) %>% #computing the means of AQI, latitude, 
                                                          #& longitude
    mutate(Month = month(Date)) %>% #adding 'Month' and 'Year' columns to the dataset by 
                                    #parsing the 'Date' column
    mutate(Year = year(Date)) 

#9. Separating AQI values for ozone and PM2.5 into two different columns
EPAair_18_19_final <-
  pivot_wider(EPAair_18_19, names_from = AQS_PARAMETER_DESC, values_from = meanAQI)

#10. Checking the dimensions of the completed dataset
dim(EPAair_18_19_final)

#11. Saving the processed dataset
write.csv(EPAair_18_19_final, row.names = FALSE, 
  file = "./Data/Processed/EPAair_O3_PM25_NC1819_Processed.csv")

```


## Generate summary tables
12. Use the split-apply-combine strategy to generate a summary data frame. Data should be grouped by site, month, and year. Generate the mean AQI values for ozone and PM2.5 for each group. Then, add a pipe to remove instances where a month and year are not available (use the function `drop_na` in your pipe). 

13. Call up the dimensions of the summary dataset.

```{r}
#12a. Using split-apply-combine strategy to generate a summary data frame
EPAair_18_19_summary <-
  EPAair_18_19_final %>% 
  group_by(Site.Name, Month, Year) %>% #group the data frame by site, month, & year
    summarise(meanAQI_Ozone = mean(Ozone), #generating mean AQI values for ozone and PM2.5
              meanAQI_PM2.5 = mean(PM2.5))
    
#12b. Adding a pipe to remove na's from columns 'meanAQI_Ozone' & 'meanAQI_PM2.5' 
        #using the 'drop_na' function
EPAair_18_19_summary <-
  EPAair_18_19_summary %>% 
    drop_na(meanAQI_Ozone) %>% 
    drop_na(meanAQI_PM2.5)

#13. Checking the dimensions of the summary dataset
dim(EPAair_18_19_summary)

```

14. Why did we use the function `drop_na` rather than `na.omit`?

> Answer: In this example, we used 'drop_na' rather than 'na.omit' because 'na.omit' is to be used for an entire data frame. Since we only want na's removed from the 'meanAQI_Ozone' and 'meanAQI_PM2.5' columns, we used the 'drop_na' function which allows us to specify columns in which we want the na's removed.