---
title: "Assignment 6: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Kelly Davidson"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
#This code chunk will tidy your knit PDF files, wrapping long code lines
#For it to work, the "formatR" package needs to be installed

#install.packages('formatR')
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE)
```

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A06_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.

## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup}
#1. Setting up my session
getwd() #checking working directory
library(tidyverse) #loading tidyverse
library(agricolae) #loading agricolae

Lake.Chemistry.raw <- read.csv("./Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv", stringsAsFactors = TRUE) #importing and naming the raw data set

Lake.Chemistry.raw$sampledate <-
  as.Date(Lake.Chemistry.raw$sampledate, format = "%m/%d/%y") #setting the 'sampledate' column  as a date object

#2. Building a ggplot theme & setting it as the default theme
A06_theme <-
  theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "dark gray"), 
  legend.position = "top")
theme_set(A06_theme)

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: During the month of July, there is no correlation or relationship between mean lake temperature and depth. In other words, the intercept and slope are equal to zero.
Ha: During the month of July, there is a relationship between mean lake temperature and depth. Further, the intercept and slope are not equal to zero.

4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4. Wrangling the raw data set with a pipe function
Lake.Chemistry.raw.wrangled <- 
  Lake.Chemistry.raw %>% 
    mutate(month = month(sampledate)) %>%  #creating a new month column from the 'sampledate' column
      filter(month == 7) %>%  #filtering to only include the month of July
        select(lakename:daynum, depth:temperature_C) %>%  #selecting specific columns as specified above
          na.omit() #omitting na's to ensure the data set only includes complete cases

#5. Plotting the wrangled data set using ggplot
tempbydepthplot <-
  ggplot(Lake.Chemistry.raw.wrangled, aes(x = depth, y = temperature_C)) + #assigning x & y values
  geom_point() + #creating a scatterplot
  geom_smooth(method = "lm") + #adding the line of best fit
  ylim(0,35) + #setting the limits of temperature along the y-axis
  ylab(expression("Temperature (C)")) + #renaming the y-axis
  xlab(expression("Depth (m)")) #renaming the x-axis
print(tempbydepthplot)

```

6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: The scatterplot of temperature by depth implies that there is a negative correlation between temperature and depth during the month of July. As temperature decreases, depth increases. In addition, the relatively equal distrubtion of points on either side of the line emphasizes the linearity of the negative correlation between temperature and depth.

7. Perform a linear regression to test the relationship and display the results

```{r linear.regression}
#7. Performing a linear regression to test the relationship between temperature & depth
tempbydepth.regression <- lm(data = Lake.Chemistry.raw.wrangled, temperature_C ~ depth)
summary(tempbydepth.regression)

```

8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: From running the linear regression, you can see that there is in fact a realtionship between temperature and depth. The p-value, 2.2e^-16, indicates whether the relationship between the two variables is statistically significant and describes this regression that is based on 9726 observation, as described by the degrees of freedom. In this scenario, the p-value is very small so we reject the null hypothesis that states there is no relationship or correlation between temperature and depth. The R-squared value of 0.7387 indicates that ~73% of the variability in temperature is explained by depth. Lastly, temperature is predicted to change -1.95 degrees celsuis for every 1 meter change in depth.

---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER.

9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9. Running an AIC to determine what explanatory variables are best suited to predict temperature
AIC.lake.chemistry <-
  lm(data = Lake.Chemistry.raw.wrangled, temperature_C ~ year4 + daynum + depth)
step(AIC.lake.chemistry)

#10. Running a multiple regression on temperature_C, year4, daynum, & depth
multiple.regression <- lm(data = Lake.Chemistry.raw.wrangled, temperature_C ~ year4 + daynum + depth)
summary(multiple.regression)

```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: 
AIC method suggests we use year4, daynum, and depth as explanatory variables to predict temperature
new r-squared value is, 0.7412 ~74% - so more of the variability in temp is explained by year4, daynum, and depth

---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12. Running an ANOVA test
Lake.Chemistry.anova <- aov(data = Lake.Chemistry.raw.wrangled, temperature_C ~ lakename)
summary(Lake.Chemistry.anova)

Lake.Chemistry.lm <- lm(data = Lake.Chemistry.raw.wrangled, temperature_C ~ lakename)
summary(Lake.Chemistry.lm)
```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: p-value, 2.2e^16, so we reject the null - there IS a significant difference in mean temperature among the lakes. R-squared value is 0.03953 - ~4% of the variability in temperature is explained by lake name

14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14. Plotting temperature by depth with separate colors for each lake
tempbydepthplot2 <-
  ggplot(Lake.Chemistry.raw.wrangled, aes(x = depth, y = temperature_C, color = lakename)) + #assigning x & y values
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  xlab("Depth (m)") +
  ylab("Temperature (C)") +
  ylim(0, 35)
print(tempbydepthplot2)

```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15. Running the Tukey HSD test
TukeyHSD(Lake.Chemistry.anova)

```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: 
same mean as Peter Lake: Ward & Paul
no lakes have a mean temperature that is statistically distinct from all the other lakes


17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: HSD test extracts groupings for pairwise relationship

18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}
#18. Running a two-sample T-test
Lake.Chemistry.Crampton.Ward <-
  Lake.Chemistry.raw.wrangled %>% 
  filter(lakename == "Crampton Lake" | lakename == "Ward Lake")

Crampton.Ward.two_sample <- t.test(Lake.Chemistry.Crampton.Ward$temperature_C ~ Lake.Chemistry.Crampton.Ward$lakename)
Crampton.Ward.two_sample

```

>Answer: The results of the t-test indicate that the means are similar, crampton lake 15.35 and Ward Lake 14.45
p-value is above 0.05 so fail to reject the null hypothesis - their means are similar
