---
title: "Assignment 09: Data Scraping"
author: "Kelly Davidson"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A09_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Check your working directory
* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Set your ggplot theme

```{r, message = FALSE}
#1. Setting up my session
getwd() #checking working directory

library(tidyverse) #loading necessary packages
library(lubridate)
library(rvest)
library(cowplot)

A10_theme <- #defining and setting my ggplot theme
  theme_light(base_size = 10) +
  theme(axis.text = element_text(color = "dark gray"), 
  legend.position = "right")
theme_set(A10_theme)

```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2021 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2021>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2. Indicating the NC DEQs Local Water Supply Planning URL to be scraped

Durham_LWSP <- read_html("https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2021")

```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PSWID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Daily Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings), with the first value being "27.6400".

```{r scrape.the.data}
#3. Scraping the data (Water System Name, PSWID, Ownership, and Maximum Daily Use) from the URL

water.system.name <- Durham_LWSP %>% 
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>% 
  html_text()
  
pswid <- Durham_LWSP %>% 
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>% 
  html_text()
  
ownership <- Durham_LWSP %>% 
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>% 
  html_text()
  
max.withdrawals.mgd <-  Durham_LWSP %>% 
  html_nodes("th~ td+ td") %>% 
  html_text()

```

4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc...

5. Create a line plot of the maximum daily withdrawals across the months for 2021

```{r create.a.dataframe.from.scraped.data}
#4. Converting the scraped data into a dataframe

Durham_LWSP_df <- data.frame("Month" = c("Jan", "May", "Sep", "Feb", "Jun", "Oct", "Mar", "Jul", "Nov", "Apr", "Aug", "Dec"),
                             "Year" = rep(2021),
                             "Max.withdrawals.mgd" = as.numeric(max.withdrawals.mgd))

Durham_LWSP_df2 <- Durham_LWSP_df %>% 
  mutate(Water.system.name = !!water.system.name,
         PSWID = !!pswid,
         Ownership =!!ownership,
         Date = my(paste(Month,"-",Year))) %>% 
  arrange(ymd(Date))

#5. Graphing a line plot of the maximum daily withdrawals for 2021

Max_daily_withdrawal_plot <- ggplot(Durham_LWSP_df2, aes(x = Date, y = Max.withdrawals.mgd)) +
  geom_line() +
  labs(title = "Maximum Daily Withdrawls in Durham in 2021",
  y = "Withdrawal (MGD)")
print(Max_daily_withdrawal_plot)

```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function using your code above that can scrape data for any PWSID and year for which the NC DEQ has data. **Be sure to modify the code to reflect the year and site (pwsid) scraped**.

```{r construct.a.scraping.function}
#6. "https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2021"

scrape.it <- function(Year, PSWID){
  the_website <- read_html(paste0("https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=", PSWID, "&year=", Year))
  
  water_system_tag <- "div+ table tr:nth-child(1) td:nth-child(2)"
  pswid_tag <- "td tr:nth-child(1) td:nth-child(5)"
  ownership_tag <- "div+ table tr:nth-child(2) td:nth-child(4)"
  max_withdrawals_tag <- "th~ td+ td"

  water_system <- the_website %>% html_nodes(water_system_tag) %>% html_text()
  pswid <- the_website %>%  html_nodes(pswid_tag) %>%  html_text()
  ownership <- the_website %>%  html_nodes(ownership_tag) %>%  html_text()
  max_withdrawals <- the_website %>%  html_nodes(max_withdrawals_tag) %>%  html_text()
  
  water_df <- data.frame("Month" = c("Jan", "May", "Sep", "Feb", "Jun", "Oct", "Mar", "Jul", "Nov", "Apr", "Aug", "Dec"),
                             "Year" = rep(Year, 12),
                             "Max_withdrawals_mgd" = as.numeric(max_withdrawals)) %>%
    mutate(Water_System = !!water_system,
         PSWID = !!pswid,
         Ownership =!!ownership,
         Date = my(paste(Month,"-",Year))) %>% 
    arrange(ymd(Date))
  
  return(water_df)
}

```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2015

```{r fetch.and.plot.Durham.2015.data}
#7. 

Durham_withdrawals_2015 <- scrape.it(2015, "03-32-010")
view(Durham_withdrawals_2015)

Durham_withdrawals_2015_plot <- ggplot(Durham_withdrawals_2015, aes(x = Date, y = Max_withdrawals_mgd)) +
  geom_line() +
  labs(title = "Maximum Daily Withdrawls in Durham in 2015",
  y = "Withdrawal (MGD)")
print(Durham_withdrawals_2015_plot)

```

8. Use the function above to extract data for Asheville (PWSID = 01-11-010) in 2015. Combine this data with the Durham data collected above and create a plot that compares Asheville's to Durham's water withdrawals.

```{r fetch.and.plot.Asheville.2015.data}
#8. 

Asheville_withdrawals_2015 <- scrape.it(2015, "01-11-010")
view(Asheville_withdrawals_2015)

Asheville_withdrawals_2015_plot <- ggplot(Asheville_withdrawals_2015, aes(x = Date, y = Max_withdrawals_mgd)) +
  geom_line() +
  labs(title = "Maximum Daily Withdrawls in Asheville in 2015",
  y = "Withdrawal (MGD)")
print(Asheville_withdrawals_2015_plot)


Ash_Durh_2015 <- rbind(Asheville_withdrawals_2015, Durham_withdrawals_2015)

Ash_Durh_2015_plot <- ggplot(Ash_Durh_2015, aes(x = Date, y = Max_withdrawals_mgd), color = Water_System) +
  geom_line()
print(Ash_Durh_2015_plot)


plot_grid(Durham_withdrawals_2015_plot, Asheville_withdrawals_2015_plot, nrow = 2, align = 'h', rel_heights = c(1.25, 1))

```


9. Use the code & function you created above to plot Asheville's max daily withdrawal by months for the years 2010 thru 2019.Add a smoothed line to the plot.

>TIP: See Section 3.2 in the "09_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs. Pipe the output of the map2() function to `bindrows()` to combine the dataframes into a single one. 

```{r fetch.and.plot.Asheville.multiyear.data}
#9


```

>Question: Just by looking at the plot (i.e. not running statistics), does Asheville have a trend in water usage over time?