---
title: "Assignment 5: Data Visualization"
author: "Kelly Davidson"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
#This code chunk will tidy your knit PDF files, wrapping long code lines
#For it to work, the "formatR" package needs to be installed

#install.packages('formatR')
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=80), tidy=TRUE)
```

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.

The completed exercise is due on Friday, Oct 14th @ 5:00pm.



## Set up your session 
1. Set up your session. Verify your working directory and load the tidyverse, 
lubridate, & cowplot packages. Upload the NTL-LTER processed data files for  
nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy 
[`NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv`] version) and 
the processed data file for the Niwot Ridge litter dataset (use the
[`NEON_NIWO_Litter_mass_trap_Processed.csv`] version). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r}
#1. Verifying working directory and loading tidyverse, lubridate, & cowplot packages
# Loading the two processed data files "Lake_nutrients_processed" and "Litter_processed)

getwd()
library(tidyverse)
library(lubridate)
library(cowplot)

Lake_nutrients_processed <- 
  read.csv("./Data/Processed/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv" ,
  stringsAsFactors = TRUE)

Litter_processed <-
  read.csv("./Data/Processed/NEON_NIWO_Litter_mass_trap_Processed.csv", 
  stringsAsFactors = TRUE)

#2. Checking the class of the date columns and changing them to the appropriate date format for both datasets

class(Lake_nutrients_processed$sampledate)
class(Litter_processed$collectDate)

Lake_nutrients_processed$sampledate <- 
  as.Date(Lake_nutrients_processed$sampledate, format = "%Y-%m-%d")
Litter_processed$collectDate <-
  as.Date(Litter_processed$collectDate, format = "%Y-%m-%d")

```
## Define your theme
3. Build a theme and set it as your default theme.

```{r}
#3. Building my theme and setting it as the default theme

A05_theme <-
  theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "dark gray"), 
  legend.position = "top")

theme_set(A05_theme)

```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best 
practices for data visualization. Ensure your theme, color palettes, axes, 
and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with 
separate aesthetics for Peter and Paul lakes. Add a line of best fit and 
color it black. Adjust your axes to hide extreme values (hint: change the 
limits using `xlim()` and/or `ylim()`).

```{r}
#4. Plotting total phosphorus by phosphate concentration with separate aesthetics for Peter Lake & Paul Lake

phosphate_plot <-
  ggplot(Lake_nutrients_processed, aes(x = tp_ug, y = po4)) +
  ylab(expression("Phosphate")) + 
  xlab(expression("Total Phosphorus")) +
  ylim(0, 50) + #adjusting y-axis to hide extreme values above 200
  geom_point(aes(shape = lakename, color = lakename)) +
  geom_smooth(method = lm, color = "black") #adding line of best fit
print(phosphate_plot)

```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and 
(c) TN, with month as the x axis and lake as a color aesthetic. Then, create 
a cowplot that combines the three graphs. Make sure that only one legend is 
present and that graph axes are aligned. 

Tip: R has a build in variable called `month.abb` that returns a list of months;
see <https://r-lang.com/month-abb-in-r-with-example>

```{r}
#5. 

Temp_boxplot <-
  ggplot(Lake_nutrients_processed, 
  aes(x = factor(month, levels = c(1:12)), y = temperature_C)) +
  xlab(expression("Month")) +
  ylab(expression("Temperature")) +
  geom_boxplot(aes(color = lakename)) +
  scale_x_discrete(drop = FALSE)
print(Temp_boxplot)

TP_boxplot <-
  ggplot(Lake_nutrients_processed,
  aes(x = factor(month, levels = c(1:12)), y = tp_ug)) +
  xlab(expression("Month")) +
  ylab(expression("Total Phosphorus")) +
  geom_boxplot(aes(color = lakename)) +
  scale_x_discrete(drop = FALSE) +
  ylim(0, 150)
print(TP_boxplot)

TN_boxplot <-
  ggplot(Lake_nutrients_processed,
  aes(x = factor(month, levels = c(1:12)), y = tn_ug)) +
  xlab(expression("Month")) +
  ylab(expression("Total Nitrogen")) +
  geom_boxplot(aes(color = lakename)) +
  scale_x_discrete(drop = FALSE)
print(TN_boxplot)


Combined_boxplot <- plot_grid(Temp_boxplot + theme(legend.position = "none"),
                            TP_boxplot + theme(legend.position = "none"),
                            TN_boxplot + theme(legend.position = "bottom"),
                            nrow = 3, rel_heights = c(3, 3, 3))
print(Combined_boxplot)

```

Question: What do you observe about the variables of interest over seasons 
and between lakes? 

> Answer: 

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the 
"Needles" functional group. Plot the dry mass of needle litter by date and 
separate by NLCD class with a color aesthetic. (no need to adjust the name 
of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into 
three facets rather than separated by color. 

```{r}
#6. 

Needles_plot <-
  ggplot(subset(Litter_processed, functionalGroup == "Needles"),
  aes(x = collectDate, y = dryMass, color = nlcdClass)) +
  xlab(expression("Date")) +
  ylab(expression("Dry Mass")) +
  geom_point()
print(Needles_plot)


#7. 

Needles_facets_plot <-
  ggplot(subset(Litter_processed, functionalGroup == "Needles"),
  aes(x = collectDate, y = dryMass)) +
  xlab(expression("Date")) +
  ylab(expression("Dry Mass")) +
  geom_point() +
  facet_wrap(vars(nlcdClass), nrow = 3)
print(Needles_facets_plot)

```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: