---
title: "Assignment 2: Coding Basics"
author: "Kelly Davidson"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics Day 1

1.  Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r}
#1. generating a sequence from 1 to 100, increasing by 4 and naming it "kelly_sequence"

kelly_sequence <- seq(1, 100, 4)

#2. computing mean and median

mean(kelly_sequence)
median(kelly_sequence)

#3. is mean > median?

isTRUE(mean(kelly_sequence) > median(kelly_sequence))

```

## Basics Day 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r}
# 5 and 6. create 3 vectors: student_names, test_scores, pass_fail

student_names <- c("Ashley" , "Brad" , "Claire" , "Dave")  #character vector
test_scores <- c(98, 72 , 49 , 61)  #numeric vector
pass_fail <- c(TRUE , TRUE , FALSE , TRUE)  #logical vector

#7. combine 3 vectors into a data frame

grades_data_frame <- data.frame(student_names, test_scores, pass_fail)

#8. assign informative titles to columns

colnames(grades_data_frame) <- c("Student Name", "Test Score", "Pass")

```

9.  QUESTION: How is this data frame different from a matrix?

> Answer: a data frame can have mixed data types (numeric, character, logical) whereas matrices can only include 1 data type

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 

11. Apply your function to the vector with test scores that you created in number 5.

```{r}

more_test_scores <- c(40, 87, 32, 95)
grading_function <- function (more_test_scores) {}

```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer:
